---
title: "Getting Started with rSHUD"
author: "rSHUD Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rSHUD}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Introduction

rSHUD is an R package toolbox for the SHUD (Simulator for Hydrologic 
Unstructured Domains) hydrological modeling system. This guide will help you
get started with building and analyzing SHUD models using modern spatial
libraries (terra and sf).

## Installation

### Install rSHUD

```{r}
# Install from GitHub
devtools::install_github("SHUD-System/rSHUD")

# Load the package
library(rSHUD)
```

### Install Dependencies

rSHUD requires modern spatial libraries:

```{r}
# Install spatial packages
install.packages(c("terra", "sf"))

# Install optional packages for advanced features
install.packages(c("gstat", "interp", "ncdf4"))
```

### System Dependencies (Linux/Mac)

On Ubuntu/Debian systems, you may need to install system libraries:

```{bash}
sudo apt-get install gdal-bin libgdal-dev
sudo apt-get install libudunits2-dev
```

## Basic Workflow

The typical rSHUD workflow consists of:

1. Load and prepare geospatial data
2. Build the model domain (mesh and river network)
3. Calculate model attributes
4. Generate input files
5. Run the SHUD model (external)
6. Analyze results

## Quick Start Example

### Load Required Libraries

```{r}
library(rSHUD)
library(terra)
library(sf)
```

### Load Example Data

rSHUD includes example datasets for demonstration:

```{r}
# Load example watershed data
data(sh)  # Shale Hills watershed

# The 'sh' object contains:
# - sh$mesh: Mesh domain
# - sh$river: River network
# - sh$att: Attributes
```

### Build a Simple Model

For a quick model build with default parameters:

```{r}
# Prepare input data (replace with your own data)
domain <- st_read("path/to/watershed.shp")
dem <- rast("path/to/dem.tif")
rivers <- st_read("path/to/rivers.shp")

# Quick model build
model <- quick_model(
  project_name = "my_watershed",
  domain = domain,
  dem = dem,
  rivers = rivers,
  output_dir = "model_output"
)

# Access results
mesh <- model$mesh
river_network <- model$river
```

## Detailed Workflow

### Step 1: Load Geospatial Data

```{r}
# Load watershed boundary (polygon)
domain <- st_read("data/watershed.shp")

# Load digital elevation model (raster)
dem <- rast("data/dem.tif")

# Load river network (lines)
rivers <- st_read("data/rivers.shp")

# Load soil data (raster)
soil <- rast("data/soil.tif")

# Load land cover (raster)
landcover <- rast("data/landcover.tif")

# Check coordinate systems
st_crs(domain)
crs(dem)
```

### Step 2: Visualize Input Data

```{r}
# Plot DEM with watershed boundary
plot(dem, main = "Digital Elevation Model")
plot(st_geometry(domain), add = TRUE, border = "red", lwd = 2)
plot(st_geometry(rivers), add = TRUE, col = "blue", lwd = 2)
```

### Step 3: Build Model with Custom Options

```{r}
# Define mesh generation options
mesh_opts <- list(
  max_area = 2e5,           # Maximum triangle area (m²)
  min_angle = 33,           # Minimum triangle angle (degrees)
  simplify_boundary = 200,  # Boundary simplification tolerance (m)
  simplify_rivers = 50      # River simplification tolerance (m)
)

# Define river processing options
river_opts <- list(
  min_length = 100  # Minimum river segment length (m)
)

# Build model
model <- auto_build_model(
  project_name = "detailed_model",
  domain = domain,
  rivers = rivers,
  dem = dem,
  soil = soil,
  landcover = landcover,
  output_dir = "model_output",
  mesh_options = mesh_opts,
  river_options = river_opts,
  aquifer_depth = 10,
  years = 2000:2010,
  verbose = TRUE
)
```

### Step 4: Examine Model Components

```{r}
# Access mesh object
mesh <- model$mesh
print(mesh)

# Get mesh statistics
n_cells <- nrow(mesh@mesh)
mesh_area <- sum(mesh@mesh$Area)
cat("Number of cells:", n_cells, "\n")
cat("Total area:", mesh_area, "m²\n")

# Access river network
river <- model$river
n_segments <- nrow(river@river)
cat("Number of river segments:", n_segments, "\n")

# View attributes
att <- model$attributes
head(att)
```

### Step 5: Visualize Model Domain

```{r}
# Convert mesh to spatial format for plotting
mesh_sf <- sp.mesh2Shape(mesh, crs = st_crs(domain))

# Plot mesh
plot(st_geometry(mesh_sf), main = "Model Mesh")
plot(st_geometry(rivers), add = TRUE, col = "blue", lwd = 2)

# Plot mesh with elevation
elev <- getElevation(mesh)
plot(mesh_sf["Elevation"], main = "Mesh Elevation")
```

### Step 6: Read and Modify Input Files

```{r}
# Read generated input files
mesh_data <- read_mesh("model_output/detailed_model.mesh")
river_data <- read_river("model_output/detailed_model.riv")
att_data <- read_att("model_output/detailed_model.att")
para_data <- read_para("model_output/detailed_model.para")

# Modify parameters
para_data$INIT_MODE <- 3  # Change initialization mode
para_data$ASCII_OUTPUT <- 1  # Enable ASCII output

# Write modified parameters
write_config(para_data, "model_output/detailed_model.para")
```

## Working with Model Results

### Read Model Output

After running the SHUD model, read the output files:

```{r}
# Read time series output
surf_depth <- read_output(
  keyword = "eleysurf",
  path = "model_output"
)

# Read outputs used by the water-balance helpers
variables <- c(
  "elevprcp", "elevetic", "elevettr", "elevetev",
  "eleqsurf", "eleqsub",
  "eleysurf", "eleyunsat", "eleygw",
  "rivqdown", "rivqsurf", "rivqsub"
)
all_output <- setNames(
  lapply(
    variables,
    read_output,
    path = "model_output"
  ),
  variables
)
```

### Visualize Results

```{r}
# Plot the first output series
plot(
  zoo::as.zoo(surf_depth[, 1]),
  main = "Surface Water Depth",
  ylab = "Depth (m)",
  xlab = "Time"
)

# Convert to raster for spatial visualization
surf_raster <- mesh_to_raster(
  data = surf_depth[1, ],  # First time step
  mesh = mesh,
  template = dem
)

plot(surf_raster, main = "Surface Water Depth at t=1")
```

### Calculate Water Balance

```{r}
# Calculate element water balance
elem_xl <- all_output[c(
  "elevprcp", "elevetic", "elevettr", "elevetev",
  "eleqsurf", "eleqsub",
  "eleysurf", "eleyunsat", "eleygw"
)]
wb_elem <- wb.ele(
  xl = elem_xl,
  fun = xts::apply.daily,
  plot = FALSE
)

# Calculate river water balance
river_xl <- all_output[c("rivqdown", "rivqsurf", "rivqsub")]
wb_river <- wb.riv(
  xl = river_xl,
  fun = xts::apply.daily,
  plot = FALSE
)

# Plot element storage change components
storage_change <- apply(wb_elem[, , c("dys", "dyu", "dyg")], 1, sum,
                        na.rm = TRUE)
plot(storage_change, type = "l", main = "Total Element Storage Change")
```

## Advanced Features

### Custom Mesh Generation

```{r}
# Create custom mesh with specific constraints
boundary <- st_read("boundary.shp")
rivers <- st_read("rivers.shp")

# Generate triangulation
tri <- shud.triangle(
  wb = boundary,
  riv = rivers,
  q = 30,      # Minimum angle
  a = 1e5      # Maximum area
)

# Create mesh object
mesh <- shud.mesh(tri, dem = dem, AqDepth = 15)
```

### River Network Processing

```{r}
# Build river network with properties
river_network <- build_river_network(
  rivers = rivers,
  dem = dem,
  min_length = 100,
  simplify = TRUE
)

# Calculate specific properties
rivers_with_order <- calc_river_properties(
  rivers = rivers,
  dem = dem,
  properties = c("order", "slope")
)

# View stream orders
table(rivers_with_order$order)
```

### Spatial Data Conversion

```{r}
# Convert mesh data to raster
elevation <- getElevation(mesh)
elev_raster <- mesh_to_raster(
  data = elevation,
  mesh = mesh,
  resolution = 100,
  method = "idw"
)

# Convert vector to raster
soil_raster <- vector_to_raster(
  vector = soil_polygons,
  template = dem,
  field = "soil_type"
)
```

### Time Series Analysis

```{r}
# Read time series data
discharge <- read_tsd("forcing/discharge.tsd")

# Convert to daily values
discharge_daily <- ts_to_daily(discharge, FUN = mean)

# Calculate flow duration curve
fdc_data <- fdc(discharge)
plot(fdc_data, main = "Flow Duration Curve")

# Plot hydrograph
QvsO(
  sim = simulated_flow,
  obs = observed_flow,
  main = "Observed vs Simulated"
)
```

## Tips and Best Practices

### Data Preparation

1. **Coordinate Systems**: Ensure all spatial data use the same CRS
2. **Resolution**: Match raster resolutions when possible
3. **Extent**: Crop data to watershed extent plus buffer
4. **Quality**: Check for missing data and outliers

```{r}
# Check and align CRS
if (!st_crs(domain) == crs(dem)) {
  domain <- st_transform(domain, crs = crs(dem))
}

# Crop rasters to domain with buffer
domain_buffered <- st_buffer(domain, dist = 2000)
dem_crop <- crop(dem, vect(domain_buffered))
```

### Mesh Quality

1. **Triangle Size**: Balance between detail and computation time
2. **Angle Quality**: Use min_angle >= 30 for stable numerics
3. **River Alignment**: Ensure rivers align with mesh edges

```{r}
# Good mesh settings for typical watershed
mesh_opts <- list(
  max_area = 2e5,    # ~14 ha triangles
  min_angle = 33,    # Good quality
  simplify_boundary = 200
)
```

### Performance

1. **Use terra/sf**: Much faster than legacy raster/sp
2. **Crop Early**: Reduce data extent before processing
3. **Simplify**: Simplify boundaries and rivers appropriately

```{r}
# Simplify complex boundaries
domain_simple <- st_simplify(domain, dTolerance = 100)

# Use appropriate resolution
dem_resample <- aggregate(dem, fact = 2)  # Reduce resolution
```

## Getting Help

### Documentation

```{r}
# Function help
?auto_build_model
?mesh_to_raster
?read_output

# Package vignettes
browseVignettes("rSHUD")
```

### Example Datasets

```{r}
# List available datasets
data(package = "rSHUD")

# Load and explore
data(sh)
str(sh)
```

### Resources

- Package website: https://www.shud.xyz/
- GitHub repository: https://github.com/SHUD-System/rSHUD
- Issue tracker: https://github.com/SHUD-System/rSHUD/issues
- SHUD model documentation: https://www.shud.xyz/Book/

## Next Steps

After completing this getting started guide:

1. Review the **Migration Guide** if upgrading from v2.1
2. Explore the **Model Building** vignette for detailed workflows
3. Check the **GIS Processing** vignette for spatial analysis
4. Read the **Hydrological Analysis** vignette for result analysis

## Complete Example Script

Here's a complete script to build and analyze a SHUD model:

```{r}
# Load libraries
library(rSHUD)
library(terra)
library(sf)

# Set working directory
setwd("my_project")

# Load data
domain <- st_read("data/watershed.shp")
dem <- rast("data/dem.tif")
rivers <- st_read("data/rivers.shp")
soil <- rast("data/soil.tif")
lc <- rast("data/landcover.tif")

# Build model
model <- auto_build_model(
  project_name = "watershed_2010",
  domain = domain,
  rivers = rivers,
  dem = dem,
  soil = soil,
  landcover = lc,
  output_dir = "model_2010",
  years = 2010:2015,
  verbose = TRUE
)

# Visualize mesh
mesh_sf <- sp.mesh2Shape(model$mesh, crs = st_crs(domain))
plot(st_geometry(mesh_sf), main = "Model Domain")

# Run SHUD model (external command)
# system("shud model_2010/watershed_2010")

# Read and analyze results
surf <- read_output(keyword = "eleysurf", path = "model_2010")
plot(
  zoo::as.zoo(surf[, 1]),
  main = "Surface Water Depth - Cell 1",
  ylab = "Depth (m)",
  xlab = "Time"
)

# Calculate water balance
wb_xl <- list(
  elevprcp = read_output(keyword = "elevprcp", path = "model_2010"),
  elevetic = read_output(keyword = "elevetic", path = "model_2010"),
  elevettr = read_output(keyword = "elevettr", path = "model_2010"),
  elevetev = read_output(keyword = "elevetev", path = "model_2010"),
  eleqsurf = read_output(keyword = "eleqsurf", path = "model_2010"),
  eleqsub = read_output(keyword = "eleqsub", path = "model_2010"),
  eleysurf = surf,
  eleyunsat = read_output(keyword = "eleyunsat", path = "model_2010"),
  eleygw = read_output(keyword = "eleygw", path = "model_2010")
)
wb <- wb.ele(xl = wb_xl, fun = xts::apply.daily, plot = FALSE)

storage_change <- apply(wb[, , c("dys", "dyu", "dyg")], 1, sum,
                        na.rm = TRUE)
plot(storage_change, type = "l", main = "Element Storage Change")
```

## Summary

This guide covered:

- Installing rSHUD and dependencies
- Basic and detailed model building workflows
- Reading and modifying input files
- Analyzing model results
- Advanced features and best practices

For more detailed information, consult the other vignettes and function
documentation.
