---
title: "GIS Processing with rSHUD"
author: "rSHUD Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GIS Processing with rSHUD}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Introduction

This vignette demonstrates GIS processing capabilities in rSHUD using modern
spatial libraries (terra and sf). It covers raster and vector operations,
coordinate transformations, spatial analysis, and data conversions.

## Prerequisites

```{r}
library(rSHUD)
library(terra)
library(sf)
```

## Raster Operations with terra

### Loading and Basic Operations

```{r}
# Load raster data
dem <- rast("data/dem.tif")

# Basic information
print(dem)
cat("Resolution:", res(dem), "\n")
cat("Extent:", as.vector(ext(dem)), "\n")
cat("CRS:", crs(dem), "\n")

# Summary statistics
global(dem, "mean", na.rm = TRUE)
global(dem, c("min", "max", "mean", "sd"), na.rm = TRUE)

# Plot
plot(dem, main = "Digital Elevation Model")
```

### Raster Algebra

```{r}
# Simple operations
dem_m <- dem  # Already in meters
dem_ft <- dem * 3.28084  # Convert to feet

# Conditional operations
slope <- terrain(dem, "slope", unit = "degrees")
steep_areas <- slope > 30  # Boolean raster

# Reclassify
rcl_matrix <- matrix(c(
  0, 100, 1,
  100, 500, 2,
  500, 1000, 3,
  1000, Inf, 4
), ncol = 3, byrow = TRUE)

elev_classes <- classify(dem, rcl_matrix)
plot(elev_classes, main = "Elevation Classes")
```

### Terrain Analysis

```{r}
# Calculate terrain derivatives
slope <- terrain(dem, "slope", unit = "degrees")
aspect <- terrain(dem, "aspect", unit = "degrees")
tpi <- terrain(dem, "TPI")  # Topographic Position Index
tri <- terrain(dem, "TRI")  # Terrain Ruggedness Index

# Flow direction and accumulation
flow_dir <- terrain(dem, "flowdir")

# Hillshade for visualization
hillshade <- shade(
  slope = terrain(dem, "slope", unit = "radians"),
  aspect = terrain(dem, "aspect", unit = "radians")
)

plot(hillshade, col = gray(0:100/100), legend = FALSE)
plot(dem, add = TRUE, alpha = 0.5)
```

### Raster Manipulation

```{r}
# Crop to extent
bbox <- ext(c(xmin = 100, xmax = 200, ymin = 100, ymax = 200))
dem_crop <- crop(dem, bbox)

# Crop to polygon
watershed <- st_read("data/watershed.shp")
dem_crop <- crop(dem, vect(watershed))
dem_mask <- mask(dem_crop, vect(watershed))

# Resample to different resolution
dem_coarse <- aggregate(dem, fact = 2, fun = "mean")
dem_fine <- disagg(dem, fact = 2, method = "bilinear")

# Reproject
dem_ll <- project(dem, "EPSG:4326", method = "bilinear")
```

### Multi-layer Rasters

```{r}
# Create raster stack
soil <- rast("data/soil.tif")
landcover <- rast("data/landcover.tif")

# Combine layers
layers <- c(dem, soil, landcover)
names(layers) <- c("elevation", "soil", "landcover")

# Access individual layers
elev_layer <- layers[[1]]
elev_layer <- layers[["elevation"]]

# Operations on all layers
layers_norm <- scale(layers)

# Extract values
pts <- data.frame(x = c(100, 200), y = c(100, 200))
values <- extract(layers, pts)
```

## Vector Operations with sf

### Loading and Basic Operations

```{r}
# Load vector data
watershed <- st_read("data/watershed.shp")
rivers <- st_read("data/rivers.shp")
stations <- st_read("data/stations.shp")

# Basic information
print(watershed)
st_geometry_type(watershed)
st_crs(watershed)
st_bbox(watershed)

# Plot
plot(st_geometry(watershed), main = "Watershed")
plot(st_geometry(rivers), add = TRUE, col = "blue")
plot(st_geometry(stations), add = TRUE, col = "red", pch = 19)
```

### Geometric Operations

```{r}
# Buffer
watershed_buf <- st_buffer(watershed, dist = 1000)
rivers_buf <- st_buffer(rivers, dist = 100)

# Union
watershed_union <- st_union(watershed)

# Intersection
rivers_in_watershed <- st_intersection(rivers, watershed)

# Difference
watershed_no_rivers <- st_difference(watershed, rivers_buf)

# Simplification
watershed_simple <- st_simplify(watershed, dTolerance = 100)
rivers_simple <- st_simplify(rivers, dTolerance = 50)

# Centroid
watershed_centroid <- st_centroid(watershed)

# Convex hull
watershed_hull <- st_convex_hull(watershed_union)
```

### Spatial Queries

```{r}
# Spatial predicates
rivers_intersect <- st_intersects(rivers, watershed)
stations_within <- st_within(stations, watershed)

# Filter based on spatial relationship
rivers_in <- rivers[lengths(st_intersects(rivers, watershed)) > 0, ]
stations_in <- stations[lengths(st_within(stations, watershed)) > 0, ]

# Distance calculations
dist_to_rivers <- st_distance(stations, rivers)
nearest_river <- apply(dist_to_rivers, 1, which.min)

# Area and length
watershed_area <- st_area(watershed)
river_lengths <- st_length(rivers)
```

### Attribute Operations

```{r}
# Select columns
rivers_subset <- rivers[, c("name", "order")]

# Filter rows
major_rivers <- rivers[rivers$order >= 3, ]

# Add new columns
rivers$length_km <- as.numeric(st_length(rivers)) / 1000

# Aggregate
rivers_by_order <- aggregate(
  rivers["length_km"],
  by = list(order = rivers$order),
  FUN = sum
)

# Join
stations$watershed_id <- st_join(stations, watershed)$id
```

## Coordinate Reference Systems

### CRS Management

```{r}
# Check CRS
st_crs(watershed)
crs(dem)

# Set CRS (if missing)
st_crs(watershed) <- 4326
crs(dem) <- "EPSG:4326"

# Transform to different CRS
watershed_utm <- st_transform(watershed, crs = 32610)  # UTM Zone 10N
dem_utm <- project(dem, "EPSG:32610")

# Transform to match another object
rivers_matched <- st_transform(rivers, crs = st_crs(watershed))
```

### Common Projections

```{r}
# Geographic (lat/lon)
watershed_ll <- st_transform(watershed, crs = 4326)

# UTM (automatically determine zone)
lon <- mean(st_coordinates(st_centroid(watershed))[, 1])
utm_zone <- crs.long2utmZone(lon)
utm_crs <- crs.long2utm(lon)
watershed_utm <- st_transform(watershed, crs = utm_crs)

# Albers Equal Area
albers_crs <- crs.Albers(
  lat1 = 29.5,
  lat2 = 45.5,
  lat0 = 37.5,
  lon0 = -96
)
watershed_albers <- st_transform(watershed, crs = albers_crs)

# Lambert Conformal Conic
lambert_crs <- crs.Lambert(
  lat1 = 33,
  lat2 = 45,
  lat0 = 39,
  lon0 = -96
)
watershed_lambert <- st_transform(watershed, crs = lambert_crs)
```

## Raster-Vector Conversions

### Mesh to Raster

```{r}
# Load mesh
mesh <- read_mesh("model/watershed.mesh")

# Get mesh data
elevation <- getElevation(mesh)
soil_type <- mesh@mesh$Soil

# Convert to raster using different methods
elev_raster_idw <- mesh_to_raster(
  data = elevation,
  mesh = mesh,
  template = dem,
  method = "idw"
)

elev_raster_linear <- mesh_to_raster(
  data = elevation,
  mesh = mesh,
  template = dem,
  method = "linear"
)

elev_raster_nearest <- mesh_to_raster(
  data = elevation,
  mesh = mesh,
  template = dem,
  method = "nearest"
)

# Compare methods
par(mfrow = c(2, 2))
plot(dem, main = "Original DEM")
plot(elev_raster_idw, main = "IDW Interpolation")
plot(elev_raster_linear, main = "Linear Interpolation")
plot(elev_raster_nearest, main = "Nearest Neighbor")
```

### Vector to Raster

```{r}
# Rasterize polygons
soil_polygons <- st_read("data/soil_polygons.shp")

soil_raster <- vector_to_raster(
  vector = soil_polygons,
  template = dem,
  field = "soil_type",
  fun = "first"
)

# Rasterize lines (rivers)
rivers_raster <- vector_to_raster(
  vector = rivers,
  template = dem,
  field = "order",
  fun = "max"
)

# Rasterize points with interpolation
stations_raster <- vector_to_raster(
  vector = stations,
  template = dem,
  field = "rainfall",
  fun = "mean"
)
```

### Raster to Vector

```{r}
# Convert raster to polygons
elev_classes <- classify(dem, rcl_matrix)
elev_polygons <- as.polygons(elev_classes)
elev_sf <- st_as_sf(elev_polygons)

# Convert raster to points
dem_points <- as.points(dem)
dem_points_sf <- st_as_sf(dem_points)

# Contour lines
contours <- as.contour(dem, levels = seq(0, 1000, 100))
contours_sf <- st_as_sf(contours)
```

## Spatial Analysis

### Watershed Delineation

```{r}
# Delineate watershed from pour point
pour_point <- st_point(c(x = 500000, y = 4000000))
pour_point_sf <- st_sfc(pour_point, crs = st_crs(dem))

watershed_delineated <- watershedDelineation(
  dem = dem,
  outlet = pour_point_sf
)

plot(dem)
plot(st_geometry(watershed_delineated), add = TRUE, border = "red", lwd = 2)
```

### River Network Extraction

```{r}
# Extract river network from DEM
flow_acc <- terrain(dem, "flowaccumulation")
threshold <- 1000  # Minimum accumulation for stream

rivers_extracted <- flow_acc > threshold
rivers_vectorized <- as.lines(rivers_extracted)

# Process river network
rivers_processed <- build_river_network(
  rivers = st_as_sf(rivers_vectorized),
  dem = dem,
  min_length = 500
)
```

### Zonal Statistics

```{r}
# Calculate statistics by zone
zones <- st_read("data/subwatersheds.shp")

# Extract raster values by polygon
zonal_stats <- extract(dem, vect(zones), fun = mean, na.rm = TRUE)
zones$mean_elev <- zonal_stats$elevation

# Multiple statistics
zonal_multi <- extract(
  dem,
  vect(zones),
  fun = function(x) c(
    mean = mean(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE)
  )
)
```

### Spatial Interpolation

```{r}
# Prepare point data
rainfall_pts <- st_read("data/rainfall_stations.shp")

# IDW interpolation
if (requireNamespace("gstat", quietly = TRUE)) {
  # Create grid
  grid <- st_as_sf(as.points(dem, values = FALSE))
  
  # Perform IDW
  idw_result <- gstat::idw(
    formula = rainfall ~ 1,
    locations = rainfall_pts,
    newdata = grid,
    idp = 2.0
  )
  
  # Convert to raster
  rainfall_raster <- rasterize(
    idw_result,
    dem,
    field = "var1.pred"
  )
}
```

## Data Quality and Validation

### Checking Data Quality

```{r}
# Check for missing values
na_count <- global(dem, "isNA")
cat("Missing values:", na_count, "\n")

# Check value ranges
value_range <- global(dem, c("min", "max"), na.rm = TRUE)
cat("Value range:", value_range$min, "to", value_range$max, "\n")

# Check for outliers
dem_values <- values(dem, na.rm = TRUE)
outliers <- which_outliers(dem_values)
cat("Outliers:", sum(outliers), "\n")

# Visualize outliers
dem_clean <- dem
dem_clean[outliers] <- NA
plot(dem_clean, main = "DEM with Outliers Removed")
```

### Spatial Coverage

```{r}
# Check raster coverage of mesh
mesh <- read_mesh("model/watershed.mesh")
mesh_sf <- mesh_to_sf(mesh, crs = st_crs(watershed))
mesh_centroids <- st_centroid(mesh_sf)
dem_values <- extract(dem, vect(mesh_centroids))[, 2]
coverage <- mean(!is.na(dem_values)) * 100
cat("DEM coverage of mesh centroids:", coverage, "%\n")

# Check vector coverage
rivers_coverage <- st_area(st_intersection(rivers, watershed)) /
                   st_area(watershed) * 100
cat("River coverage:", rivers_coverage, "%\n")
```

### Alignment Checks

```{r}
# Check if extents overlap
dem_ext <- ext(dem)
watershed_ext <- st_bbox(watershed)

overlaps <- !(
  dem_ext$xmax < watershed_ext["xmin"] ||
  dem_ext$xmin > watershed_ext["xmax"] ||
  dem_ext$ymax < watershed_ext["ymin"] ||
  dem_ext$ymin > watershed_ext["ymax"]
)

cat("Extents overlap:", overlaps, "\n")

# Check CRS compatibility
crs_match <- compatible_crs(crs(dem), st_crs(watershed))
cat("CRS compatible:", crs_match, "\n")
```

## Performance Optimization

### Efficient Raster Processing

```{r}
# Process in chunks for large rasters
process_large_raster <- function(r, fun) {
  # Process by blocks
  result <- app(r, fun, cores = 4)  # Use multiple cores
  return(result)
}

# Use appropriate data types
dem_int <- as.int(dem * 100)  # Store as integer with scaling

# Crop before processing
dem_crop <- crop(dem, vect(watershed))
slope <- terrain(dem_crop, "slope")  # Faster than full extent
```

### Efficient Vector Processing

```{r}
# Simplify before processing
rivers_simple <- st_simplify(rivers, dTolerance = 50)

# Use spatial indexing
rivers_indexed <- st_make_valid(rivers)

# Batch operations
watersheds <- st_read("data/watersheds.shp")
areas <- st_area(watersheds)  # Vectorized operation
```

## Visualization

### Basic Plotting

```{r}
# Raster visualization
plot(dem, main = "Elevation", col = terrain.colors(100))
contour(dem, add = TRUE, levels = seq(0, 1000, 100))

# Vector visualization
plot(st_geometry(watershed), main = "Watershed", border = "red", lwd = 2)
plot(st_geometry(rivers), add = TRUE, col = "blue", lwd = 1)
plot(st_geometry(stations), add = TRUE, col = "black", pch = 19)
```

### Advanced Visualization

```{r}
# Hillshade with overlay
hillshade <- shade(
  slope = terrain(dem, "slope", unit = "radians"),
  aspect = terrain(dem, "aspect", unit = "radians")
)

plot(hillshade, col = gray(0:100/100), legend = FALSE, main = "Hillshade DEM")
plot(dem, add = TRUE, alpha = 0.5, col = terrain.colors(100))
plot(st_geometry(watershed), add = TRUE, border = "red", lwd = 2)

# Multi-panel plots
par(mfrow = c(2, 2))
plot(dem, main = "Elevation")
plot(terrain(dem, "slope"), main = "Slope")
plot(terrain(dem, "aspect"), main = "Aspect")
plot(terrain(dem, "TPI"), main = "TPI")
```

### Interactive Visualization

```{r}
# Using mapview (if available)
if (requireNamespace("mapview", quietly = TRUE)) {
  mapview::mapview(watershed) +
    mapview::mapview(rivers, color = "blue") +
    mapview::mapview(stations, col.regions = "red")
}
```

## Complete GIS Workflow Example

```{r}
# Complete GIS processing workflow
process_watershed_data <- function(
  dem_file,
  watershed_file,
  rivers_file,
  output_dir = "processed"
) {
  
  # Create output directory
  dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
  
  # Load data
  cat("Loading data...\n")
  dem <- rast(dem_file)
  watershed <- st_read(watershed_file)
  rivers <- st_read(rivers_file)
  
  # Align CRS
  cat("Aligning coordinate systems...\n")
  if (!compatible_crs(crs(dem), st_crs(watershed))) {
    watershed <- st_transform(watershed, crs = crs(dem))
  }
  if (!compatible_crs(crs(dem), st_crs(rivers))) {
    rivers <- st_transform(rivers, crs = crs(dem))
  }
  
  # Crop and mask DEM
  cat("Cropping DEM...\n")
  watershed_buf <- st_buffer(watershed, dist = 2000)
  dem_crop <- crop(dem, vect(watershed_buf))
  dem_mask <- mask(dem_crop, vect(watershed))
  
  # Calculate terrain derivatives
  cat("Calculating terrain derivatives...\n")
  slope <- terrain(dem_mask, "slope", unit = "degrees")
  aspect <- terrain(dem_mask, "aspect", unit = "degrees")
  
  # Process rivers
  cat("Processing river network...\n")
  rivers_clip <- st_intersection(rivers, watershed)
  rivers_simple <- st_simplify(rivers_clip, dTolerance = 50)
  
  river_network <- build_river_network(
    rivers = rivers_simple,
    dem = dem_mask,
    min_length = 100
  )
  
  # Save results
  cat("Saving results...\n")
  writeRaster(dem_mask, file.path(output_dir, "dem.tif"), overwrite = TRUE)
  writeRaster(slope, file.path(output_dir, "slope.tif"), overwrite = TRUE)
  writeRaster(aspect, file.path(output_dir, "aspect.tif"), overwrite = TRUE)
  
  st_write(watershed, file.path(output_dir, "watershed.shp"),
           delete_dsn = TRUE)
  st_write(river_network$network, file.path(output_dir, "rivers.shp"),
           delete_dsn = TRUE)
  
  cat("Processing complete!\n")
  
  return(list(
    dem = dem_mask,
    slope = slope,
    aspect = aspect,
    watershed = watershed,
    rivers = river_network
  ))
}

# Run the workflow
results <- process_watershed_data(
  dem_file = "data/dem.tif",
  watershed_file = "data/watershed.shp",
  rivers_file = "data/rivers.shp",
  output_dir = "processed_data"
)
```

## Summary

This vignette covered:

- Raster operations with terra
- Vector operations with sf
- Coordinate reference system management
- Raster-vector conversions
- Spatial analysis techniques
- Data quality and validation
- Performance optimization
- Visualization methods
- Complete GIS workflow

For more information, see the other vignettes and function documentation.
