---
title: "Hydrological Analysis with rSHUD"
author: "rSHUD Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hydrological Analysis with rSHUD}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Introduction

This vignette demonstrates hydrological analysis capabilities in rSHUD,
including reading model output, calculating water balance, analyzing time
series, and evaluating model performance.

## Prerequisites

```{r}
library(rSHUD)
library(terra)
library(sf)
library(xts)
library(hydroGOF)
```

## Reading Model Output

### Output File Structure

SHUD model outputs are stored in binary or ASCII format:

- `<project>.eleysurf.dat`: Surface water depth
- `<project>.eleyunsat.dat`: Unsaturated zone storage
- `<project>.eleygw.dat`: Groundwater storage
- `<project>.rivystage.dat`: River stage
- `<project>.rivqdown.dat`: River discharge
- `<project>.elevprcp.dat`: Precipitation
- `<project>.elevinfil.dat`: Infiltration rate
- `<project>.elevrech.dat`: Groundwater recharge
- `<project>.eleveta.dat`: Actual evapotranspiration
- `<project>.elevetp.dat`: Potential evapotranspiration
- `<project>.elevetic.dat`: Canopy interception evaporation
- `<project>.elevettr.dat`: Transpiration
- `<project>.elevetev.dat`: Soil evaporation
- `<project>.eleqsurf.dat`: Element surface flow
- `<project>.eleqsub.dat`: Element subsurface flow
- `<project>.rivqsurf.dat`: Surface inflow to river
- `<project>.rivqsub.dat`: Groundwater inflow to river

### Reading Single Variable

```{r}
# Read surface water output
surf <- read_output(
  path = "model_output",
  keyword = "eleysurf"
)

# Check dimensions
dim(surf)  # [time_steps, n_cells]

# View first few time steps
head(surf[, 1:5])

# Summary statistics
summary(surf[1, ])  # First time step
```

### Reading Multiple Variables

```{r}
# Read outputs used in this analysis
variables <- c(
  "eleysurf", "eleyunsat", "eleygw",
  "elevprcp", "eleveta", "elevetp",
  "elevetic", "elevettr", "elevetev",
  "eleqsurf", "eleqsub",
  "rivystage", "rivqdown", "rivqsurf", "rivqsub"
)
output <- setNames(
  lapply(
    variables,
    read_output,
    path = "model_output"
  ),
  variables
)

# Access individual variables
surf <- output$eleysurf
unsat <- output$eleyunsat
gw <- output$eleygw
river_stage <- output$rivystage
river_discharge <- output$rivqdown
```

### Reading with Time Range

```{r}
# Read specific time period
output_subset <- read_output(
  path = "model_output",
  keyword = "eleysurf"
)

# Convert to daily values after reading
output_daily <- read_output(
  path = "model_output",
  keyword = "eleysurf"
)
output_daily <- ts_to_daily(output_daily)
```

## Water Balance Analysis

### Element Water Balance

```{r}
# Calculate water balance for mesh elements
elem_xl <- output[c(
  "elevprcp", "elevetic", "elevettr", "elevetev",
  "eleqsurf", "eleqsub",
  "eleysurf", "eleyunsat", "eleygw"
)]
wb_elem <- wb.ele(
  xl = elem_xl,
  fun = xts::apply.daily,
  plot = FALSE
)

# Components
storage_surf <- wb_elem[, , "dys"]    # Surface storage change
storage_unsat <- wb_elem[, , "dyu"]   # Unsaturated zone storage change
storage_gw <- wb_elem[, , "dyg"]      # Groundwater storage change
storage_total <- apply(wb_elem[, , c("dys", "dyu", "dyg")], 1, sum,
                       na.rm = TRUE)

# Plot total storage over time
plot(storage_total, type = "l", main = "Total Element Storage Change",
     ylab = "Storage change (m)", xlab = "Time")
```

### River Water Balance

```{r}
# Calculate river water balance
river_xl <- output[c("rivqdown", "rivqsurf", "rivqsub")]
wb_riv <- wb.riv(
  xl = river_xl,
  fun = xts::apply.daily,
  plot = FALSE
)

# Components
river_outflow <- wb_riv[, "Qout"]     # Outlet discharge
river_surface_in <- wb_riv[, "Qin_sf"]  # Surface inflow
river_gw_in <- wb_riv[, "Qin_gw"]     # Groundwater inflow

# Plot discharge at outlet
outlet_id <- ncol(river_discharge)
plot(river_discharge[, outlet_id],
     main = "Outlet Discharge",
     ylab = "Discharge (m³/s)", xlab = "Time")
```

### Watershed Water Balance

```{r}
# Calculate watershed-scale water balance
watershed_xl <- output[c(
  "elevprcp", "eleveta", "elevetp", "rivqdown",
  "eleysurf", "eleyunsat", "eleygw", "rivystage"
)]
wb_all <- wb.all(
  xl = watershed_xl,
  ic = read_ic(shud.filein(inpath = "model_input")["md.ic"]),
  fun = xts::apply.monthly,
  plot = FALSE
)

# Water balance components
storage_residual <- wb_all[, "D_pqe"]
precipitation <- wb_all[, "P"]
evapotranspiration <- wb_all[, "AET"]
runoff <- wb_all[, "Q"]
potential_et <- wb_all[, "PET"]

# Plot water balance
plot(storage_residual, main = "Watershed Water Balance Residual",
     ylab = "Depth (mm)", xlab = "Time")
```

### Storage Change Analysis

```{r}
# Calculate storage change
surface_storage_change <- DeltaS(surf)
unsat_storage_change <- DeltaS(unsat)
gw_storage_change <- DeltaS(gw)
delta_storage <- surface_storage_change + unsat_storage_change +
  gw_storage_change

# Plot storage change by element
plot(delta_storage, main = "Storage Change",
     ylab = "ΔS (m)", xlab = "Element")
abline(h = 0, col = "red", lty = 2)

# Cumulative storage change
cumulative_ds <- cumsum(delta_storage)
plot(cumulative_ds, main = "Cumulative Storage Change",
     ylab = "Cumulative ΔS (m)", xlab = "Element")
```

## Time Series Analysis

### Converting to Time Series

```{r}
# Convert to xts time series
surf_ts <- ts_to_df(surf)

# Convert to daily values
surf_daily <- ts_to_daily(surf, FUN = mean)

# Extract specific cells
cell_ids <- c(1, 100, 500)
surf_subset <- surf[, cell_ids]
```

### Aggregation

```{r}
# Daily aggregation
surf_daily_mean <- ts_to_daily(surf, FUN = mean)
surf_daily_max <- ts_to_daily(surf, FUN = max)
surf_daily_min <- ts_to_daily(surf, FUN = min)

# Monthly aggregation
surf_monthly <- apply.monthly(surf_daily_mean, mean)

# Annual aggregation
surf_annual <- apply.yearly(surf_daily_mean, mean)

# Plot aggregated data
plot(surf_daily_mean[, 1], main = "Daily Mean Surface Water",
     ylab = "Depth (m)", xlab = "Time")
lines(surf_monthly[, 1], col = "red", lwd = 2)
```

### Statistical Analysis

```{r}
# Calculate statistics
mean_surf <- apply(surf, 2, mean)
sd_surf <- apply(surf, 2, sd)
max_surf <- apply(surf, 2, max)
min_surf <- apply(surf, 2, min)

# Coefficient of variation
cv_surf <- sd_surf / mean_surf

# Percentiles
q25_surf <- apply(surf, 2, quantile, probs = 0.25)
q75_surf <- apply(surf, 2, quantile, probs = 0.75)

# Plot spatial distribution of statistics
mesh_sf <- sp.mesh2Shape(mesh)
mesh_sf$mean_surf <- mean_surf
plot(mesh_sf["mean_surf"], main = "Mean Surface Water Depth")
```

## Hydrograph Analysis

### Plotting Hydrographs

```{r}
# Simple hydrograph
plot(
  zoo::as.zoo(river_discharge[, outlet_id]),
  main = "Outlet Hydrograph",
  ylab = "Discharge (m³/s)",
  xlab = "Time"
)

# Multiple hydrographs
sel_ids <- c(1, 10, 20, outlet_id)
sel_flow <- river_discharge[, sel_ids]
matplot(
  zoo::index(sel_flow),
  zoo::coredata(sel_flow),
  type = "l",
  lty = 1,
  col = seq_along(sel_ids),
  xlab = "Time",
  ylab = "Discharge (m³/s)",
  main = "River Discharge at Multiple Locations"
)
legend(
  "topright",
  legend = c("Segment 1", "Segment 10", "Segment 20", "Outlet"),
  col = seq_along(sel_ids),
  lty = 1,
  bty = "n"
)
```

### Observed vs Simulated

```{r}
# Load observed data
obs_discharge <- read_tsd("data/observed_discharge.tsd")[[1]]

# Align time periods
common_time <- index(obs_discharge)[index(obs_discharge) %in%
                                    index(river_discharge)]
obs_aligned <- obs_discharge[common_time]
sim_aligned <- river_discharge[common_time, outlet_id]

# Plot comparison
QvsO(
  sim = sim_aligned,
  obs = obs_aligned,
  main = "Observed vs Simulated Discharge"
)

# Add statistics to plot
gof_stats <- gof(sim = sim_aligned, obs = obs_aligned)
legend("topright", legend = c(
  paste("NSE:", round(gof_stats["NSE"], 3)),
  paste("R²:", round(gof_stats["R2"], 3)),
  paste("PBIAS:", round(gof_stats["PBIAS"], 1), "%")
))
```

### Flow Duration Curve

```{r}
# Calculate flow duration curve
fdc_data <- fdc(river_discharge[, outlet_id])

# Plot FDC
fdc_data +
  ggplot2::labs(
    title = "Flow Duration Curve",
    x = "Exceedance Probability (%)",
    y = "Discharge (m³/s)"
  ) +
  ggplot2::scale_y_log10()

# Compare observed and simulated FDC
fdc(cbind(Observed = obs_discharge, Simulated = sim_aligned)) +
  ggplot2::labs(
    title = "Flow Duration Curves",
    x = "Exceedance Probability (%)",
    y = "Discharge (m³/s)"
  ) +
  ggplot2::scale_y_log10()
```

### Peak Flow Analysis

```{r}
# Identify peaks
peaks <- which(diff(sign(diff(river_discharge[, outlet_id]))) == -2) + 1
peak_values <- river_discharge[peaks, outlet_id]
peak_times <- index(river_discharge)[peaks]

# Plot with peaks highlighted
plot(river_discharge[, outlet_id], main = "Discharge with Peaks")
points(peak_times, peak_values, col = "red", pch = 19)

# Peak statistics
cat("Number of peaks:", length(peaks), "\n")
cat("Mean peak discharge:", mean(peak_values), "m³/s\n")
cat("Max peak discharge:", max(peak_values), "m³/s\n")
```

## Model Performance Evaluation

### Goodness-of-Fit Statistics

```{r}
# Calculate comprehensive GOF statistics
gof_stats <- gof(
  sim = sim_aligned,
  obs = obs_aligned,
  na.rm = TRUE
)

print(gof_stats)

# Individual statistics
nse <- NSE(sim = sim_aligned, obs = obs_aligned)
rmse <- rmse(sim = sim_aligned, obs = obs_aligned)
pbias <- pbias(sim = sim_aligned, obs = obs_aligned)
r2 <- cor(sim_aligned, obs_aligned)^2

cat("Nash-Sutcliffe Efficiency:", nse, "\n")
cat("Root Mean Square Error:", rmse, "\n")
cat("Percent Bias:", pbias, "%\n")
cat("R-squared:", r2, "\n")
```

### Visual Performance Assessment

```{r}
# Scatter plot
plot(obs_aligned, sim_aligned,
     main = "Observed vs Simulated",
     xlab = "Observed Discharge (m³/s)",
     ylab = "Simulated Discharge (m³/s)",
     pch = 19, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red", lwd = 2)  # 1:1 line

# Residual plot
residuals <- sim_aligned - obs_aligned
plot(obs_aligned, residuals,
     main = "Residual Plot",
     xlab = "Observed Discharge (m³/s)",
     ylab = "Residuals (m³/s)",
     pch = 19, col = rgb(0, 0, 0, 0.3))
abline(h = 0, col = "red", lwd = 2)

# Q-Q plot
qqnorm(residuals, main = "Normal Q-Q Plot of Residuals")
qqline(residuals, col = "red", lwd = 2)
```

### Performance by Flow Regime

```{r}
# Separate by flow regime
q_median <- median(obs_aligned)

# Low flows
low_flow_idx <- obs_aligned < q_median
nse_low <- NSE(sim_aligned[low_flow_idx], obs_aligned[low_flow_idx])

# High flows
high_flow_idx <- obs_aligned >= q_median
nse_high <- NSE(sim_aligned[high_flow_idx], obs_aligned[high_flow_idx])

cat("NSE (low flows):", nse_low, "\n")
cat("NSE (high flows):", nse_high, "\n")

# Performance by season
obs_df <- data.frame(
  time = index(obs_aligned),
  obs = coredata(obs_aligned),
  sim = coredata(sim_aligned)
)
obs_df$season <- quarters(obs_df$time)

# Calculate NSE by season
library(dplyr)
season_stats <- obs_df %>%
  group_by(season) %>%
  summarise(
    NSE = NSE(sim, obs),
    RMSE = rmse(sim, obs),
    PBIAS = pbias(sim, obs)
  )

print(season_stats)
```

## Spatial Analysis

### Spatial Distribution of Variables

```{r}
# Convert time step to raster
time_step <- 100
surf_raster <- mesh_to_raster(
  data = surf[time_step, ],
  mesh = mesh,
  template = dem,
  method = "idw"
)

plot(surf_raster, main = paste("Surface Water at t =", time_step))

# Create animation of spatial distribution
if (requireNamespace("animation", quietly = TRUE)) {
  animation::saveGIF({
    for (t in seq(1, nrow(surf), by = 10)) {
      r <- mesh_to_raster(surf[t, ], mesh, dem, method = "idw")
      plot(r, main = paste("Time step:", t))
    }
  }, movie.name = "surf_animation.gif")
}
```

### Spatial Statistics

```{r}
# Calculate spatial statistics for each time step
spatial_mean <- apply(surf, 1, mean)
spatial_sd <- apply(surf, 1, sd)
spatial_max <- apply(surf, 1, max)
spatial_min <- apply(surf, 1, min)

# Plot spatial statistics over time
par(mfrow = c(2, 2))
plot(spatial_mean, main = "Spatial Mean", ylab = "Depth (m)")
plot(spatial_sd, main = "Spatial SD", ylab = "Depth (m)")
plot(spatial_max, main = "Spatial Max", ylab = "Depth (m)")
plot(spatial_min, main = "Spatial Min", ylab = "Depth (m)")
```

### Hotspot Analysis

```{r}
# Identify cells with high variability
cv_cells <- apply(surf, 2, function(x) sd(x) / mean(x))

# Map high variability cells
mesh_sf$cv <- cv_cells
high_var_cells <- mesh_sf[cv_cells > quantile(cv_cells, 0.9), ]

plot(st_geometry(mesh_sf), col = "lightgray", border = NA)
plot(st_geometry(high_var_cells), col = "red", add = TRUE)
title("High Variability Cells (top 10%)")
```

## Evapotranspiration Analysis

### Reading ET Output

```{r}
# Read ET components
et_total <- read_output(keyword = "eleveta", path = "model_output")
et_canopy <- read_output(keyword = "elevetic", path = "model_output")
et_transpiration <- read_output(keyword = "elevettr", path = "model_output")
et_soil <- read_output(keyword = "elevetev", path = "model_output")

# Calculate watershed-average ET
et_mean <- apply(et_total, 1, mean)

# Plot ET time series
plot(et_mean, main = "Watershed-Average Evapotranspiration",
     ylab = "ET (mm/day)", xlab = "Time")
```

### ET Partitioning

```{r}
# Calculate ET components as fraction of total
et_canopy_frac <- apply(et_canopy, 1, sum) / apply(et_total, 1, sum)
et_transp_frac <- apply(et_transpiration, 1, sum) / apply(et_total, 1, sum)
et_soil_frac <- apply(et_soil, 1, sum) / apply(et_total, 1, sum)

# Plot ET partitioning
et_components <- data.frame(
  time = index(et_mean),
  canopy = et_canopy_frac,
  transpiration = et_transp_frac,
  soil = et_soil_frac
)

# Stacked area plot
library(ggplot2)
ggplot(et_components, aes(x = time)) +
  geom_area(aes(y = canopy, fill = "Canopy")) +
  geom_area(aes(y = canopy + transpiration, fill = "Transpiration")) +
  geom_area(aes(y = canopy + transpiration + soil, fill = "Soil")) +
  labs(title = "ET Partitioning", y = "Fraction", x = "Time") +
  scale_fill_manual(values = c("green", "blue", "brown"))
```

### PET Calculation

```{r}
# Calculate potential ET using Penman-Monteith
forcing_data <- read_tsd("forcing/meteo.tsd")[[1]]

pet <- PET_PM(
  Wind = forcing_data$wind,
  Temp = forcing_data$temp,
  RH = forcing_data$rh,
  RadNet = forcing_data$radiation,
  Press = NULL,
  Elevation = mean(getElevation(mesh))
)

# Compare PET and actual ET
plot(pet, main = "PET vs Actual ET",
     ylab = "ET (mm/day)", xlab = "Time", col = "red")
lines(et_mean, col = "blue")
legend("topright", legend = c("PET", "Actual ET"),
       col = c("red", "blue"), lty = 1)

# Calculate ET/PET ratio
et_pet_ratio <- et_mean / pet
plot(et_pet_ratio, main = "ET/PET Ratio",
     ylab = "Ratio", xlab = "Time")
abline(h = 1, col = "red", lty = 2)
```

## Groundwater Analysis

### Groundwater Levels

```{r}
# Calculate groundwater depth below surface
gw_depth <- mesh@mesh$zmax - gw

# Plot groundwater depth
gw_depth_mean <- apply(gw_depth, 1, mean)
plot(gw_depth_mean, main = "Mean Groundwater Depth",
     ylab = "Depth (m)", xlab = "Time")

# Spatial distribution of groundwater
gw_raster <- mesh_to_raster(
  data = gw[100, ],
  mesh = mesh,
  template = dem,
  method = "idw"
)
plot(gw_raster, main = "Groundwater Level")
```

### Baseflow Separation

```{r}
# Simple baseflow separation
# (Implement recursive digital filter or other method)
baseflow_index <- function(discharge, alpha = 0.925) {
  n <- length(discharge)
  baseflow <- numeric(n)
  baseflow[1] <- discharge[1]
  
  for (i in 2:n) {
    baseflow[i] <- alpha * baseflow[i-1] +
                   (1 - alpha) * (discharge[i] + discharge[i-1]) / 2
    baseflow[i] <- min(baseflow[i], discharge[i])
  }
  
  return(baseflow)
}

# Calculate baseflow
discharge <- river_discharge[, outlet_id]
baseflow <- baseflow_index(discharge)
quickflow <- discharge - baseflow

# Plot components
plot(discharge, main = "Flow Components",
     ylab = "Discharge (m³/s)", xlab = "Time")
lines(baseflow, col = "blue", lwd = 2)
lines(quickflow, col = "red", lwd = 2)
legend("topright", legend = c("Total", "Baseflow", "Quickflow"),
       col = c("black", "blue", "red"), lty = 1, lwd = c(1, 2, 2))

# Baseflow index
bfi <- sum(baseflow) / sum(discharge)
cat("Baseflow Index:", bfi, "\n")
```

## Complete Analysis Workflow

```{r}
# Complete hydrological analysis workflow
analyze_shud_output <- function(
  output_dir,
  mesh_file,
  input_dir = dirname(mesh_file),
  obs_file = NULL,
  start_date = NULL,
  end_date = NULL
) {
  
  # Load mesh
  cat("Loading mesh...\n")
  mesh <- read_mesh(mesh_file)
  
  # Read model output
  cat("Reading model output...\n")
  variables <- c(
    "eleysurf", "eleyunsat", "eleygw",
    "elevprcp", "eleveta", "elevetp",
    "rivystage", "rivqdown"
  )
  output <- setNames(
    lapply(
      variables,
      read_output,
      path = output_dir
    ),
    variables
  )
  
  # Calculate water balance
  cat("Calculating water balance...\n")
  watershed_xl <- output[c(
    "elevprcp", "eleveta", "elevetp", "rivqdown",
    "eleysurf", "eleyunsat", "eleygw", "rivystage"
  )]
  wb <- wb.all(
    xl = watershed_xl,
    ic = read_ic(shud.filein(inpath = input_dir)["md.ic"]),
    fun = xts::apply.monthly,
    plot = FALSE
  )
  
  # Plot water balance
  plot(wb[, "D_pqe"], main = "Water Balance Residual")
  
  # Analyze discharge
  cat("Analyzing discharge...\n")
  outlet_id <- ncol(output$rivqdown)
  discharge <- output$rivqdown[, outlet_id]
  
  # Flow duration curve
  fdc_data <- fdc(discharge)
  plot(fdc_data, main = "Flow Duration Curve", log = "y")
  
  # If observed data available, calculate GOF
  if (!is.null(obs_file)) {
    cat("Evaluating model performance...\n")
    obs <- read_tsd(obs_file)[[1]]
    
    # Align time periods
    common_time <- zoo::index(obs)[zoo::index(obs) %in% zoo::index(discharge)]
    obs_aligned <- obs[common_time]
    sim_aligned <- discharge[common_time]
    
    # Calculate statistics
    gof_stats <- gof(sim = sim_aligned, obs = obs_aligned)
    print(gof_stats)
    
    # Plot comparison
    QvsO(sim = sim_aligned, obs = obs_aligned)
  }
  
  # Spatial analysis
  cat("Creating spatial maps...\n")
  surf_mean <- apply(output$eleysurf, 2, mean)
  surf_raster <- mesh_to_raster(
    data = surf_mean,
    mesh = mesh,
    method = "idw"
  )
  plot(surf_raster, main = "Mean Surface Water Depth")
  
  cat("Analysis complete!\n")
  
  return(list(
    water_balance = wb,
    discharge = discharge,
    gof = if (!is.null(obs_file)) gof_stats else NULL
  ))
}

# Run analysis
results <- analyze_shud_output(
  output_dir = "model_output",
  mesh_file = "model/watershed.mesh",
  input_dir = "model_input",
  obs_file = "data/observed_discharge.tsd",
  start_date = "2010-01-01",
  end_date = "2010-12-31"
)
```

## Summary

This vignette covered:

- Reading SHUD model output
- Water balance analysis
- Time series analysis and aggregation
- Hydrograph analysis and flow duration curves
- Model performance evaluation
- Spatial analysis of model results
- Evapotranspiration analysis
- Groundwater and baseflow analysis
- Complete analysis workflow

For more information, see the other vignettes and function documentation.
