---
title: "Migration Guide: rSHUD v2.1 to v2.2"
author: "rSHUD Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migration Guide: rSHUD v2.1 to v2.2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

rSHUD v2.2 represents a major modernization of the package, transitioning from
legacy spatial libraries (raster, sp, rgeos) to modern alternatives (terra, sf).
This guide will help you migrate your existing code to work with the new version.

## Key Changes

### 1. Spatial Library Migration

**raster → terra**

The `raster` package has been completely replaced with `terra`, which offers
significantly better performance and a more consistent API.

```{r}
# OLD (v2.1 and earlier)
library(raster)
dem <- raster("dem.tif")
dem_crop <- crop(dem, extent(boundary))
slope <- terrain(dem_crop, "slope")

# NEW (v2.2)
library(terra)
dem <- rast("dem.tif")
dem_crop <- crop(dem, boundary)
slope <- terrain(dem_crop, "slope")
```

**sp → sf**

The `sp` package has been replaced with `sf` for vector data operations.

```{r}
# OLD (v2.1 and earlier)
library(sp)
library(rgdal)
watershed <- readOGR("watershed.shp")
watershed_buf <- gBuffer(watershed, width = 1000)
area <- gArea(watershed_buf)

# NEW (v2.2)
library(sf)
watershed <- st_read("watershed.shp")
watershed_buf <- st_buffer(watershed, dist = 1000)
area <- st_area(watershed_buf)
```

### 2. Function Renaming

Many functions have been renamed to follow modern R conventions (snake_case).
The old names are still available but deprecated and will show warnings.

#### I/O Functions

| Old Name | New Name | Status |
|----------|----------|--------|
| `readmesh()` | `read_mesh()` | Deprecated |
| `readriv()` | `read_river()` | Deprecated |
| `readatt()` | `read_att()` | Deprecated |
| `readpara()` | `read_para()` | Deprecated |
| `readcalib()` | `read_calib()` | Deprecated |
| `readic()` | `read_ic()` | Deprecated |
| `readlai()` | `read_lai()` | Deprecated |
| `readlc()` | `read_lc()` | Deprecated |
| `readsoil()` | `read_soil()` | Deprecated |
| `readgeol()` | `read_geol()` | Deprecated |
| `read.tsd()` | `read_tsd()` | Deprecated |
| `readout()` | `read_output()` | Deprecated |

#### GIS Functions

| Old Name | New Name | Status |
|----------|----------|--------|
| `MeshData2Raster()` | `mesh_to_raster()` | Deprecated |
| `sp2raster()` | `vector_to_raster()` | Deprecated |
| `ProjectCoordinate()` | `project_coords()` | Deprecated |
| `sp.RiverOrder()` | `calc_river_order()` | Deprecated |
| `sp.RiverPath()` | `calc_river_path()` | Deprecated |
| `sp.RiverDown()` | `calc_river_downstream()` | Deprecated |

#### Main Interface

| Old Name | New Name | Status |
|----------|----------|--------|
| `autoBuildModel()` | `auto_build_model()` | Deprecated |

### 3. Parameter Changes

Some function parameters have been renamed or changed:

```{r}
# mesh_to_raster (formerly MeshData2Raster)
# OLD
r <- MeshData2Raster(x = data, pm = mesh, rmask = template, proj = crs)

# NEW
r <- mesh_to_raster(data = data, mesh = mesh, template = template, crs = crs)

# vector_to_raster (formerly sp2raster)
# OLD
r <- sp2raster(sp = polygons, mask = template)

# NEW
r <- vector_to_raster(vector = polygons, template = template)
```

### 4. Return Type Changes

Functions now return modern spatial objects:

```{r}
# OLD - Returns RasterLayer
dem <- raster("dem.tif")
class(dem)  # "RasterLayer"

# NEW - Returns SpatRaster
dem <- terra::rast("dem.tif")
class(dem)  # "SpatRaster"

# OLD - Returns SpatialPolygonsDataFrame
watershed <- readOGR("watershed.shp")
class(watershed)  # "SpatialPolygonsDataFrame"

# NEW - Returns sf object
watershed <- sf::st_read("watershed.shp")
class(watershed)  # "sf" "data.frame"
```

## Migration Steps

### Step 1: Update Package Installation

```{r}
# Install updated rSHUD
devtools::install_github("SHUD-System/rSHUD")

# Ensure modern spatial packages are installed
install.packages(c("terra", "sf"))
```

### Step 2: Update Data Loading

Replace all `raster` and `sp` data loading with `terra` and `sf`:

```{r}
# OLD
library(raster)
library(rgdal)
dem <- raster("data/dem.tif")
soil <- raster("data/soil.tif")
watershed <- readOGR("data/watershed.shp")
rivers <- readOGR("data/rivers.shp")

# NEW
library(terra)
library(sf)
dem <- rast("data/dem.tif")
soil <- rast("data/soil.tif")
watershed <- st_read("data/watershed.shp")
rivers <- st_read("data/rivers.shp")
```

### Step 3: Update Function Calls

Replace deprecated function names with new ones:

```{r}
# OLD
mesh <- readmesh("model.mesh")
river <- readriv("model.riv")
att <- readatt("model.att")

# NEW
mesh <- read_mesh("model.mesh")
river <- read_river("model.riv")
att <- read_att("model.att")
```

### Step 4: Update Model Building Workflow

The main model building function has been significantly updated:

```{r}
# OLD (v2.1)
library(raster)
library(rgdal)

# Load data with old libraries
dem <- raster("dem.tif")
soil <- raster("soil.tif")
lc <- raster("landcover.tif")
wbd <- readOGR("boundary.shp")
rivers <- readOGR("rivers.shp")

# Build model
model <- autoBuildModel(
  indata = list(
    dem = dem,
    soil = soil,
    lc = lc,
    wbd = wbd,
    rivers = rivers
  ),
  prjname = "my_watershed",
  outpath = "output"
)

# NEW (v2.2)
library(terra)
library(sf)

# Load data with modern libraries
dem <- rast("dem.tif")
soil <- rast("soil.tif")
lc <- rast("landcover.tif")
domain <- st_read("boundary.shp")
rivers <- st_read("rivers.shp")

# Build model with new interface
model <- auto_build_model(
  project_name = "my_watershed",
  domain = domain,
  rivers = rivers,
  dem = dem,
  soil = soil,
  landcover = lc,
  output_dir = "output",
  years = 2000:2010,
  verbose = TRUE
)
```

## Common Migration Patterns

### Pattern 1: Raster Operations

```{r}
# OLD
library(raster)
r1 <- raster("file1.tif")
r2 <- raster("file2.tif")
r_crop <- crop(r1, extent(r2))
r_resample <- resample(r1, r2)
r_stack <- stack(r1, r2)
values(r1)[values(r1) < 0] <- 0

# NEW
library(terra)
r1 <- rast("file1.tif")
r2 <- rast("file2.tif")
r_crop <- crop(r1, r2)
r_resample <- resample(r1, r2)
r_stack <- c(r1, r2)
r1[r1 < 0] <- 0
```

### Pattern 2: Vector Operations

```{r}
# OLD
library(sp)
library(rgeos)
poly <- readOGR("polygon.shp")
poly_buf <- gBuffer(poly, width = 100)
poly_union <- gUnaryUnion(poly)
area <- gArea(poly)

# NEW
library(sf)
poly <- st_read("polygon.shp")
poly_buf <- st_buffer(poly, dist = 100)
poly_union <- st_union(poly)
area <- st_area(poly)
```

### Pattern 3: Coordinate Projection

```{r}
# OLD
library(sp)
library(rgdal)
poly_ll <- spTransform(poly, CRS("+proj=longlat +datum=WGS84"))

# NEW
library(sf)
poly_ll <- st_transform(poly, crs = 4326)
# or
poly_ll <- st_transform(poly, crs = "+proj=longlat +datum=WGS84")
```

### Pattern 4: Raster-Vector Conversion

```{r}
# OLD
library(raster)
library(sp)
r <- raster("data.tif")
poly <- rasterToPolygons(r)
r2 <- rasterize(poly, r, field = "value")

# NEW
library(terra)
r <- rast("data.tif")
poly <- as.polygons(r)
r2 <- rasterize(poly, r, field = "value")
```

## Troubleshooting

### Error: "Legacy raster/sp objects are not supported"

If you see this error, it means you're passing old-style spatial objects to
rSHUD functions. Convert them first:

```{r}
# Convert raster to terra
if (inherits(my_raster, "Raster")) {
  my_raster <- terra::rast(my_raster)
}

# Convert sp to sf
if (inherits(my_spatial, "Spatial")) {
  my_spatial <- sf::st_as_sf(my_spatial)
}
```

### Warning: "Function is deprecated"

Update your function calls to use the new names. The old names will be removed
in a future version.

```{r}
# Instead of
mesh <- readmesh("model.mesh")  # Warning: deprecated

# Use
mesh <- read_mesh("model.mesh")  # No warning
```

### Performance Issues

If you experience performance issues, ensure you're using the latest versions
of terra and sf:

```{r}
update.packages(c("terra", "sf"))
```

## Benefits of Migration

Migrating to v2.2 provides several benefits:

1. **Performance**: terra is 2-10x faster than raster for most operations
2. **Memory**: Better memory management for large datasets
3. **Consistency**: More consistent API across spatial operations
4. **Maintenance**: Active development and support
5. **Features**: Access to new spatial analysis capabilities

## Getting Help

If you encounter issues during migration:

1. Check the function documentation: `?function_name`
2. Review the inst/MIGRATION_GUIDE.md file in the package
3. Visit the package website: https://www.shud.xyz/
4. Report issues: https://github.com/SHUD-System/rSHUD/issues

## Complete Example

Here's a complete before/after example:

```{r}
# ============================================================================
# OLD CODE (v2.1)
# ============================================================================
library(raster)
library(rgdal)
library(rSHUD)

# Load data
dem <- raster("data/dem.tif")
soil <- raster("data/soil.tif")
lc <- raster("data/landcover.tif")
wbd <- readOGR("data/boundary.shp")
rivers <- readOGR("data/rivers.shp")

# Build model
model <- autoBuildModel(
  indata = list(dem = dem, soil = soil, lc = lc, wbd = wbd, rivers = rivers),
  prjname = "watershed",
  outpath = "output"
)

# Read results
mesh <- readmesh("output/watershed.mesh")
river <- readriv("output/watershed.riv")

# Convert to raster
elev_raster <- MeshData2Raster(
  x = getElevation(mesh),
  pm = mesh,
  rmask = dem
)

# ============================================================================
# NEW CODE (v2.2)
# ============================================================================
library(terra)
library(sf)
library(rSHUD)

# Load data
dem <- rast("data/dem.tif")
soil <- rast("data/soil.tif")
lc <- rast("data/landcover.tif")
domain <- st_read("data/boundary.shp")
rivers <- st_read("data/rivers.shp")

# Build model
model <- auto_build_model(
  project_name = "watershed",
  domain = domain,
  rivers = rivers,
  dem = dem,
  soil = soil,
  landcover = lc,
  output_dir = "output",
  years = 2000:2010
)

# Read results
mesh <- read_mesh("output/watershed.mesh")
river <- read_river("output/watershed.riv")

# Convert to raster
elev_raster <- mesh_to_raster(
  data = getElevation(mesh),
  mesh = mesh,
  template = dem
)
```

## Summary

The migration from v2.1 to v2.2 primarily involves:

1. Replacing `raster` with `terra`
2. Replacing `sp` with `sf`
3. Updating function names to snake_case
4. Adjusting parameter names where changed

While these changes require updates to existing code, they provide significant
improvements in performance, maintainability, and access to modern spatial
analysis capabilities.
