---
title: "Model Building with rSHUD"
author: "rSHUD Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Building with rSHUD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette provides detailed guidance on building SHUD hydrological models
using rSHUD. It covers the complete workflow from data preparation through
mesh generation, attribute calculation, and input file creation.

## Prerequisites

```{r}
library(rSHUD)
library(terra)
library(sf)
```

## Model Building Overview

The SHUD model building process consists of several key steps:

1. **Data Preparation**: Load and preprocess geospatial data
2. **Domain Definition**: Define watershed boundary and river network
3. **Mesh Generation**: Create unstructured triangular mesh
4. **Attribute Calculation**: Extract soil, geology, land cover properties
5. **River Processing**: Build river network and calculate properties
6. **Initial Conditions**: Set initial state variables
7. **Parameters**: Configure model parameters and calibration
8. **Time Series**: Generate forcing data time series

## Data Preparation

### Required Data

Minimum required data:
- **Watershed boundary**: Polygon defining model domain
- **Digital Elevation Model (DEM)**: Raster with elevation data

Optional but recommended:
- **River network**: Polylines for streams and rivers
- **Soil data**: Raster with soil types or properties
- **Geology data**: Raster with geological units
- **Land cover**: Raster with vegetation/land use classes
- **Forcing locations**: Points for meteorological forcing

### Loading Data

```{r}
# Load watershed boundary
domain <- st_read("data/watershed.shp")

# Load DEM
dem <- rast("data/dem.tif")

# Load optional data
rivers <- st_read("data/rivers.shp")
soil <- rast("data/soil.tif")
geology <- rast("data/geology.tif")
landcover <- rast("data/landcover.tif")
```

### Data Quality Checks

```{r}
# Check coordinate systems
cat("Domain CRS:", st_crs(domain)$input, "\n")
cat("DEM CRS:", crs(dem, describe = TRUE)$name, "\n")

# Check for missing values
cat("DEM NA count:", global(is.na(dem), "sum")$sum, "\n")

# Check extents
cat("Domain extent:", st_bbox(domain), "\n")
cat("DEM extent:", ext(dem), "\n")

# Visualize
plot(dem, main = "Input Data")
plot(st_geometry(domain), add = TRUE, border = "red", lwd = 2)
if (!is.null(rivers)) {
  plot(st_geometry(rivers), add = TRUE, col = "blue", lwd = 1)
}
```

### Data Preprocessing

```{r}
# Align coordinate systems
if (st_crs(domain) != crs(dem)) {
  cat("Reprojecting domain to match DEM...\n")
  domain <- st_transform(domain, crs = crs(dem))
}

# Crop and buffer DEM
buffer_dist <- 2000  # 2 km buffer
domain_buffered <- st_buffer(domain, dist = buffer_dist)
dem_crop <- crop(dem, vect(domain_buffered))

# Fill sinks in DEM (if needed)
# dem_filled <- fill_sinks(dem_crop)

# Resample rasters to common resolution (if needed)
target_res <- 30  # 30m resolution
if (res(dem_crop)[1] != target_res) {
  dem_resample <- aggregate(dem_crop, fact = target_res / res(dem_crop)[1])
}
```

## Mesh Generation

### Understanding Mesh Parameters

The mesh quality is controlled by several parameters:

- **max_area**: Maximum triangle area (m²)
  - Smaller = more detail, longer computation
  - Typical: 1e5 to 5e5 m² (10-50 hectares)
  
- **min_angle**: Minimum triangle angle (degrees)
  - Higher = better quality, fewer triangles
  - Typical: 30-35 degrees
  - Must be < 34 degrees for Delaunay triangulation
  
- **simplify_boundary**: Tolerance for boundary simplification (m)
  - Reduces complexity of watershed outline
  - Typical: 100-500 m
  
- **simplify_rivers**: Tolerance for river simplification (m)
  - Reduces river network complexity
  - Typical: 0-100 m

### Basic Mesh Generation

```{r}
# Simple mesh with default parameters
mesh_opts <- list(
  max_area = 2e5,
  min_angle = 33,
  simplify_boundary = 200,
  simplify_rivers = 0
)

# Generate triangulation
tri <- shud.triangle(
  wb = domain,
  riv = rivers,
  q = mesh_opts$min_angle,
  a = mesh_opts$max_area
)

# Create mesh object
mesh <- shud.mesh(tri, dem = dem_crop, AqDepth = 10)

# Check mesh statistics
cat("Number of triangles:", nrow(mesh@mesh), "\n")
cat("Number of nodes:", nrow(mesh@point), "\n")
cat("Total area:", sum(mesh@mesh$Area), "m²\n")
cat("Mean triangle area:", mean(mesh@mesh$Area), "m²\n")
```

### Advanced Mesh Generation

```{r}
# Simplify boundary for coarser mesh
domain_simple <- st_simplify(domain, dTolerance = mesh_opts$simplify_boundary)

# Simplify rivers
if (!is.null(rivers) && mesh_opts$simplify_rivers > 0) {
  rivers_simple <- st_simplify(rivers, dTolerance = mesh_opts$simplify_rivers)
  
  # Remove short segments
  lengths <- as.numeric(st_length(rivers_simple))
  rivers_simple <- rivers_simple[lengths >= 100, ]
} else {
  rivers_simple <- rivers
}

# Generate mesh with simplified inputs
tri <- shud.triangle(
  wb = domain_simple,
  riv = rivers_simple,
  q = mesh_opts$min_angle,
  a = mesh_opts$max_area
)

mesh <- shud.mesh(tri, dem = dem_crop, AqDepth = 10)
```

### Mesh Quality Assessment

```{r}
# Calculate triangle angles
angles <- mesh@mesh[, c("Ang1", "Ang2", "Ang3")]
min_angles <- apply(angles, 1, min)
max_angles <- apply(angles, 1, max)

# Plot angle distribution
hist(min_angles, breaks = 30, main = "Minimum Triangle Angles",
     xlab = "Angle (degrees)", col = "lightblue")
abline(v = mesh_opts$min_angle, col = "red", lwd = 2, lty = 2)

# Check for poor quality triangles
poor_quality <- sum(min_angles < 20)
cat("Triangles with min angle < 20°:", poor_quality, "\n")

# Plot mesh
mesh_sf <- sp.mesh2Shape(mesh, crs = st_crs(domain))
plot(st_geometry(mesh_sf), main = "Generated Mesh", border = "gray")
plot(st_geometry(domain), add = TRUE, border = "red", lwd = 2)
```

### Adaptive Mesh Refinement

```{r}
# Create variable resolution mesh
# Finer mesh near rivers, coarser elsewhere

# Define refinement zones
river_buffer <- st_buffer(rivers, dist = 500)

# Generate mesh with constraints
# (This requires custom implementation)
# For now, use uniform mesh with appropriate max_area
```

## Attribute Calculation

### Soil and Geology Attributes

```{r}
# Extract soil types to mesh cells
soil_attr <- shud.att(
  tri = tri,
  r.soil = soil,
  r.geol = geology
)

# View attribute summary
summary(soil_attr)

# Check for missing values
cat("Cells with missing soil:", sum(is.na(soil_attr$Soil)), "\n")
cat("Cells with missing geology:", sum(is.na(soil_attr$Geol)), "\n")

# Apply pedotransfer functions
soil_params <- PTF.soil(soil_attr$Soil)
geol_params <- PTF.geol(soil_attr$Geol)

# Combine attributes
mesh_att <- cbind(soil_attr, soil_params, geol_params)
```

### Land Cover Attributes

```{r}
# Extract land cover
lc_attr <- shud.att(
  tri = tri,
  r.lc = landcover
)

# Map land cover to vegetation parameters
# (Depends on land cover classification system)
if (all(lc_attr$LC >= 0 & lc_attr$LC <= 100)) {
  # NLCD classification
  veg_params <- LaiRf.NLCD(lc_attr$LC)
} else {
  # GLC classification
  veg_params <- LaiRf.GLC(lc_attr$LC)
}

# Add to attributes
mesh_att <- cbind(mesh_att, lc_attr, veg_params)
```

### Forcing Data Assignment

```{r}
# Assign forcing locations to mesh cells
if (!is.null(forcing_sites)) {
  forc_attr <- shud.att(
    tri = tri,
    r.forc = forcing_sites
  )
  
  mesh_att$Forc <- forc_attr$Forc
} else {
  # Use single forcing for all cells
  mesh_att$Forc <- 1
}
```

### Complete Attribute Table

```{r}
# Create complete attribute table
att_complete <- data.frame(
  INDEX = 1:nrow(mesh@mesh),
  SOIL = mesh_att$Soil,
  GEOL = mesh_att$Geol,
  LC = mesh_att$LC,
  FORC = mesh_att$Forc,
  LAI = mesh_att$LAI,
  RL = mesh_att$RL,
  # Add soil hydraulic properties
  ThetaS = soil_params$ThetaS,
  ThetaR = soil_params$ThetaR,
  Ksat = soil_params$Ksat,
  # Add other parameters...
  stringsAsFactors = FALSE
)

# Write attributes
write_df(att_complete, "model/watershed.att")
```

## River Network Processing

### River Network Building

```{r}
# Build complete river network with all properties
river_network <- build_river_network(
  rivers = rivers,
  dem = dem_crop,
  min_length = 100,
  simplify = TRUE,
  tolerance = 50
)

# Access components
river_sf <- river_network$network
river_order <- river_network$order
river_downstream <- river_network$downstream
river_props <- river_network$properties

# Summary statistics
cat("Total river segments:", nrow(river_sf), "\n")
cat("Stream orders:", paste(unique(river_order), collapse = ", "), "\n")
cat("Number of outlets:", sum(river_downstream == -1), "\n")
```

### River Properties

```{r}
# Calculate individual properties
rivers_with_props <- calc_river_properties(
  rivers = rivers,
  dem = dem_crop,
  properties = c("order", "downstream", "length", "slope")
)

# View properties
head(rivers_with_props)

# Plot by stream order
plot(rivers_with_props["order"], main = "Stream Order",
     pal = terrain.colors(max(rivers_with_props$order)))
```

### River-Mesh Integration

```{r}
# Cut rivers with mesh triangles
mesh_sf <- sp.mesh2Shape(mesh, crs = st_crs(domain))
river_segments <- shud.rivseg(mesh_sf, river_sf)

# Check connectivity
cat("River-mesh segments:", nrow(river_segments), "\n")
cat("Cells with rivers:", length(unique(river_segments$iEle)), "\n")

# Write river segment file
write_df(river_segments, "model/watershed.rivseg")
```

### River Width Estimation

```{r}
# Estimate river width from drainage area
watershed_area <- as.numeric(st_area(domain)) * 1e-6  # km²

river_widths <- calc_river_width_from_area(
  area = watershed_area,
  order = river_order
)

# Add to river properties
river_props$Width <- river_widths
```

## Initial Conditions

### Setting Initial Conditions

```{r}
# Generate default initial conditions
n_cells <- nrow(mesh@mesh)
n_rivers <- nrow(river_network$network)

ic <- shud.ic(
  nelem = n_cells,
  nriver = n_rivers,
  AqD = 10  # Aquifer depth
)

# Customize initial conditions
# Set initial water table depth
ic$Canopy <- 0  # No canopy storage
ic$Snow <- 0    # No snow
ic$Surf <- 0    # No surface water
ic$Unsat <- 0.3 * mesh@mesh$zmax  # 30% of soil depth
ic$Gw <- 0.5 * 10  # 50% of aquifer depth

# Set river initial conditions
ic$RivStg <- 0.1  # 10 cm stage
ic$RivGw <- 0.5   # 50 cm groundwater

# Write initial conditions
write_ic(ic, "model/watershed.ic")
```

## Model Parameters

### Parameter Configuration

```{r}
# Generate default parameters
n_days <- 365 * 10  # 10 years
para <- shud.para(nday = n_days)

# Customize parameters
para$INIT_MODE <- 3  # Relaxation mode
para$ASCII_OUTPUT <- 1  # Enable ASCII output
para$SPINUP_DAY <- 365  # 1 year spinup

# Time step settings
para$DT <- 60  # 60 second time step
para$START <- "2000-01-01 00:00"
para$END <- "2010-12-31 23:59"

# Output intervals
para$SURF_STEP <- 1440  # Daily surface output
para$UNSAT_STEP <- 1440  # Daily unsaturated zone output
para$GW_STEP <- 1440  # Daily groundwater output
para$RIVER_STEP <- 1440  # Daily river output

# Write parameters
write_config(para, "model/watershed.para")
```

### Calibration Parameters

```{r}
# Generate default calibration parameters
calib <- shud.calib()

# Adjust calibration factors
calib$GEOL_KSATH <- 1.0  # Horizontal hydraulic conductivity
calib$GEOL_KSATV <- 1.0  # Vertical hydraulic conductivity
calib$GEOL_KMACSATH <- 1.0  # Macropore conductivity
calib$GEOL_KMACSATV <- 1.0

calib$SOIL_KINF <- 1.0  # Infiltration
calib$SOIL_KSAT <- 1.0  # Soil saturated conductivity
calib$SOIL_KMACV <- 1.0  # Soil macropore conductivity

calib$LC_ROUGH <- 1.0  # Surface roughness
calib$LC_VEGFRAC <- 1.0  # Vegetation fraction

# Write calibration parameters
write_config(calib, "model/watershed.calib")
```

## Time Series Data

### LAI and Roughness Length

```{r}
# Generate LAI and roughness length time series
years <- 2000:2010
lc_values <- unique(mesh_att$LC)

lai_rl <- fun.lairl(lc_values, years = years)

# View time series
plot(lai_rl$LAI, main = "Leaf Area Index")
plot(lai_rl$RL, main = "Roughness Length")

# Write time series
write_tsd(lai_rl$LAI, "model/watershed.lai")
write_tsd(lai_rl$RL, "model/watershed.rl")
```

### Melt Factor

```{r}
# Generate melt factor time series
mf <- MeltFactor(years = years)

plot(mf, main = "Melt Factor")

write_tsd(mf, "model/watershed.mf")
```

### Forcing Data

```{r}
# List forcing files
forcing_files <- c(
  "forcing/prcp.txt",
  "forcing/temp.txt",
  "forcing/rh.txt",
  "forcing/wind.txt",
  "forcing/solar.txt"
)

# Write forcing file list
write_forc(
  forcing_files,
  file = "model/watershed.forc",
  startdate = "20000101"
)
```

## Complete Workflow Example

```{r}
# Complete model building workflow
build_complete_model <- function(
  project_name,
  domain_file,
  dem_file,
  rivers_file = NULL,
  soil_file = NULL,
  geology_file = NULL,
  landcover_file = NULL,
  output_dir = "model",
  years = 2000:2010
) {
  
  # Load data
  cat("Loading data...\n")
  domain <- st_read(domain_file)
  dem <- rast(dem_file)
  
  rivers <- if (!is.null(rivers_file)) st_read(rivers_file) else NULL
  soil <- if (!is.null(soil_file)) rast(soil_file) else NULL
  geology <- if (!is.null(geology_file)) rast(geology_file) else NULL
  landcover <- if (!is.null(landcover_file)) rast(landcover_file) else NULL
  
  # Build model
  cat("Building model...\n")
  model <- auto_build_model(
    project_name = project_name,
    domain = domain,
    rivers = rivers,
    dem = dem,
    soil = soil,
    geology = geology,
    landcover = landcover,
    output_dir = output_dir,
    years = years,
    verbose = TRUE
  )
  
  cat("Model building complete!\n")
  cat("Output directory:", output_dir, "\n")
  
  return(model)
}

# Use the function
model <- build_complete_model(
  project_name = "my_watershed",
  domain_file = "data/watershed.shp",
  dem_file = "data/dem.tif",
  rivers_file = "data/rivers.shp",
  soil_file = "data/soil.tif",
  landcover_file = "data/landcover.tif",
  output_dir = "model_output"
)
```

## Troubleshooting

### Common Issues

**Issue: Mesh generation fails**
```{r}
# Try reducing max_area or increasing min_angle
# Check for self-intersecting boundaries
domain_valid <- st_is_valid(domain)
if (!domain_valid) {
  domain <- st_make_valid(domain)
}
```

**Issue: Missing attribute values**
```{r}
# Check raster coverage
coverage <- ForcingCoverage(mesh, dem)
cat("DEM coverage:", coverage, "%\n")

# Fill missing values
att$Soil[is.na(att$Soil)] <- median(att$Soil, na.rm = TRUE)
```

**Issue: River network disconnected**
```{r}
# Check for multiple outlets
outlets <- get_river_outlets(river_network$network)
cat("Number of outlets:", length(outlets), "\n")

# Manually connect if needed
```

## Summary

This vignette covered:

- Data preparation and quality checks
- Mesh generation with quality control
- Attribute calculation and assignment
- River network processing and integration
- Initial conditions and parameters
- Time series data generation
- Complete workflow implementation

For more information, see the other vignettes and function documentation.
